---
title: "Soil Diagnostic Features & Characteristics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Soil Diagnostic Features & Characteristics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SoilTaxonomy)
```

Data from the latest version of the Keys to Soil Taxonomy related to diagnostic features and characteristics are stored in the package dataset `ST_features`. 

You can load the full supporting data directly with:

```r
data(ST_features, package = "SoilTaxonomy")
```

A helpful wrapper function is provided, `get_ST_features()` that helps with obtaining information of interest to specific features or groups of features.

We will inspect the "Mollic Epipedon" information to see the structure of the `ST_features` _data.frame_.

We can use `get_ST_features(name=...)` to get features by name (case-insensitive)

```{r}
str(get_ST_features(name = "mollic epipedon"))
```

Columns included in the `ST_features`, and usable for filtering as arguments to `get_ST_features()` include: 

 - `group` - Diagnostic feature or characteristic group name
 - `name` - Diagnostic feature or characteristic name
 - `chapter` - Chapter number in Keys to Soil Taxonomy (latest edition)
 - `page` - Page number in Keys to Soil Taxonomy (latest edition)
 - `description` - Narrative description (may not be present for all features)
 - `criteria` - "Key" to identifying the feature; logical criteria separated by and/or

Say we are interested in finding criteria to distinguish the epipedons (surface diagnostic horizons).

We can use the `group=...` argument to get all of the "surface" features as a data.frame.

```{r}
epipedons <- get_ST_features(group = "surface")
```

There are `r nrow(epipedons)` features in this result set.

```{r}
str(epipedons)
```

We can then select specific items by name, such as the `"description"` and `"criteria"`:

```{r}
subset(epipedons, name == "Anthropic Epipedon")$description
```

```{r}
subset(epipedons, name == "Anthropic Epipedon")$criteria
```

If we want to compare the basic definitions in Chapter 2 with the features in Chapter 3, we can subset ST_features using the chapter (or page!) numbers.

```{r}
get_ST_features(chapter = 2)$name
```

```{r}
get_ST_features(chapter = 3)$name
```
