---
title: "Taxon Letter Codes in Soil Taxonomy"
output: rmarkdown::html_vignette
author: Andrew Brown
date: 2021-04-18
vignette: >
  %\VignetteIndexEntry{Taxon Letter Codes in Soil Taxonomy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

There are four different "levels" of the hierarchy in Soil Taxonomy that are represented by letter codes: 

 - Soil Order 
 - Suborder
 - Great Group
 - Subgroup

In the _SoilTaxonomy_ package the `level` argument is used by some functions to specify output for a target level within the hierarchy. Other functions determine `level` by comparison against known taxa or codes. This vignette covers the basics of how taxon letter code conversion to and from taxonomic names is implemented.

```{r setup}
library(SoilTaxonomy)
```

## `taxon_to_taxon_code()`

`taxon_to_taxon_code()` converts a taxon name (Soil Order, Suborder, Great Group, or Subgroup) to a letter code that corresponds to the logical position of that taxon in the Keys to Soil Taxonomy. 

_Gelisols_ are the first Soil Order to key out and are given letter code "A"

```{r}
taxon_to_taxon_code("gelisols")
```

The number of letters in a taxon code corresponds to the `level` of that taxon. 
_Histels_ are the first Suborder to key out in the _Gelisols_ key (**A**), so they are given two letter code "AA"

```{r}
taxon_to_taxon_code("histels")
```

For each "step" in each key, the letter codes are "incremented" by one. 

_Glacistels_ are the <u>second</u> Great Group in the _Histels_ key (**AA**), so they have the three letter code "AAB".  

```{r}
taxon_to_taxon_code("glacistels")
```

_Typic_ subgroups, by convention, are the last subgroup to key out in a Great Group. 

```{r}
taxon_to_taxon_code("typic glacistels")
```

Since _Typic Glacistels_ have code `"AABC"` we can infer that there are three taxa in the _Glacistels_ key with codes `"AABA"`, `"AABB"` and `"AABC"`

This follows for Great Groups with many more subgroups. In case a Great Group has more than 26 subgroups within it, a fifth lowercase letter code is used to "extend" the ability to increment the code beyond 26. 

An example of where this is needed is in the _Haploxerolls_ key where the _Typic_ subgroup has code `"IFFZh"`.

```{r}
taxon_to_taxon_code("typic haploxerolls")
```

From this code we infer that the _Haploxerolls_ key has $26+8=34$ subgroups corresponding to the range from `IFFA` to `IFFZ` _plus_ `IFFZa` to `IFFZh`.

## `taxon_code_to_taxon()`

We can use a vector of letter codes to do the inverse operation with `taxon_code_to_taxon()`.

Above we determined the _Glacistels_ Key contains three taxa with codes `"AABA"`, `"AABB"` and `"AABC"`. Let's convert those codes to taxon names.

```{r}
taxon_code_to_taxon(c("AABA", "AABB", "AABC"))
```

## `taxon_to_level()`

We can infer from the length of the four-letter codes that all of the above are subgroup-level taxa. `taxon_to_level()` confirms this.

```{r}
taxon_to_level(c("Hemic Glacistels","Sapric Glacistels","Typic Glacistels"))
```

`taxon_to_level()` can also identify a fifth (lower-level) _family_ tier (`level="family"`). Soil family differentia are not handled in the Order to Subgroup keys. Family names are defined by concatenating comma-separated class names on to the subgroup. Classes used in family names are determined by specific keys and apply variably depending on the subgroup-level taxonomy.

For instance, the soil family `"Fine, mixed, semiactive, mesic Ultic Haploxeralfs"` includes a particle-size class (`"fine"`), a mineralogy class (`"mixed"`), a cation exchange capacity (CEC) activity class (`"semiactive"`) and a temperature class (`"mesic"`)

```{r}
taxon_to_level("Fine, mixed, semiactive, mesic Ultic Haploxeralfs")
```

## `getTaxonAtLevel()`

A wrapper method around taxon letter code functionality is `getTaxonAtLevel()`. 

Say that you have family-level taxon above and you want to determine the taxonomy at a higher (less detailed) level. 
You can determine what to remove (family and subgroup-level modifiers) to get the Great Group using `getTaxonAtLevel(level="greatgroup")`

```{r}
getTaxonAtLevel("Fine, mixed, semiactive, mesic Ultic Haploxeralfs", level = "greatgroup")
```

If you request a more-detailed taxonomic level than what you start with, you will get an `NA` result.

For example, we request the subgroup from suborder (`"Folists"`) level taxon name which is undefined.

```{r}
getTaxonAtLevel("Folists", level = "subgroup")
```

## `getParentTaxa()`

Another wrapper method around taxon letter code functionality is `getParentTaxa()`. This function will enumerate the tiers above a particular taxon.

```{r}
getParentTaxa("Fine, mixed, semiactive, mesic Ultic Haploxeralfs")
```

You can alternately specify `code` argument instead of `taxon`.

```{r}
getParentTaxa(code = "BAB")
```

And converting the internally used taxon codes to taxon names can be disabled with `convert = FALSE`. This may be useful for certain applications.

```{r}
getParentTaxa(code = c("BAA","BAB"), convert = FALSE)
```

## `decompose_taxon_code()`

For more general cases `decompose_taxon_code()` might be useful. This is a function used by many of the above methods that returns a nested list result containing the letter code hierarchy.

```{r}
decompose_taxon_code(c("BAA","BAB"))
```

## `preceding_taxon_codes()` and `relative_taxon_code_position()`

Other unctions useful for comparing relative positions within Keys, or the number of "steps" that it takes to reach a particular taxon, are `preceding_taxon_codes()` and `relative_taxon_code_positon()`.

`preceding_taxon_codes()` returns a list of vectors containing all preceding codes. 

For example, the `AA` suborder key precedes `AB`. And within the `AB` key `ABA` and `ABB` precede `ABC`.

```{r}
preceding_taxon_codes("ABC")
```

`relative_taxon_code_position()` counts how many taxa key out before a taxon plus $1$ (to get _the_ taxon position).

```{r}
relative_taxon_code_position(c("A","AA","AAA","AAAA",
                               "AB","AAB","ABA","ABC",
                               "B","BA","BAA","BAB",
                               "BBA","BBB","BBC"))
```

